---
title: "Get started with R - MPH1 introduction course"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to read data, recode, filter observations in a data frame and perfom basic statiscal summary.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
tutorial_options(
  exercise.timelimit = 60,
  # A simple checker function that just returns the message in the check chunk
  exercise.checker = function(check_code, ...) {
    list(
      message = eval(parse(text = check_code)),
      correct = logical(0),
      type = "info",
      location = "append"
    )
  }
)

options(tutorial.storage = list(
  # save an arbitrary R object "data" to storage
  save_object = function(tutorial_id, tutorial_version, user_id, object_id, data) {
  },
  # retreive a single R object from storage
  get_object = function(tutorial_id, tutorial_version, user_id, object_id) { 
    NULL 
  },
  # retreive a list of all R objects stored
  get_objects = function(tutorial_id, tutorial_version, user_id) { 
    list() 
  },
  # remove all stored R objects
  remove_all_objects = function(tutorial_id, tutorial_version, user_id) {
  }))
  
knitr::opts_chunk$set(error = TRUE, tutorial.data_dir="data")
load("../../data/sleepapnea.Rdata")
```

## Welcome

This is an **interactive version of the MPH1 computer labs** with the statistical software *R*. **Your objective is to learn the basic of R** while running online examples and trying some exercises. The tutorial uses the Sleep apnea data set, presented in the next section.

### 

In this tutorial, you will learn how to:

-   read a dataset stored on your computer (as a CSV file)
-   explore the data format and the different type of variables
-   compute descriptive statistic
-   display plots
-   compute univariate statistical test
-   compute simple linear and logistic regression

Every R code chunk is link to a R engine and, when you hit the *Run code* button, the code is interpreted.

```{r example, exercise = TRUE, exercise.eval = FALSE}
```

For the examples, we invite you to run the code and for the exercises we invite you to go beyond the example by adding arguments or completing with other R functions.

[As companion websites, we invite you to look at:]{.underline}

<https://nolwenn.github.io/ebook-M1-biostat/>

<https://nolwenn.github.io/software-M1-biostat/>

## The Sleep Apnea Study

All examples and proposed exercices are base on the Sleep Apnea study.

The present study deals with the sleep apnea problem. Sleep apnea is a sleep disorder characterizing by abnormal pauses in breathing during sleep. Each pause in breathing is called apnea.

Sleep apnea can affect anyone at any age. Well known risk factors for sleep apnea include being a male, overweight, age over 40, stress, and having a family history of sleep apnea. If left untreated, sleep apnea can result in a growing number of health problems, including high blood pressure, stroke, heart failure, irregular heartbeats, heart attacks, diabetes, fatigue, road accidents, and depression. Indeed, as some patients can stop breathing very often, for instance, 30 times in an hour, i.e. every 2 minutes, it means that they never get restful (or restorative) sleep. Therefore, they don't feel good during the day and their activity during the night cause stress for the heart. Prevalence of sleep apnea is estimated to be between 1 and 10% in the general population.

The main objective of this study is to explore in a French sample of population the determinants of the sleep apnea. The population was defined as the population hospitalized in the Rennes University hospital and aged more than 18 years old. A sampling procedure has been performed to select a random sample. Several individual characteristics have been collected. Their sleep has been monitored for a night (6 hours) and the number of pauses of breathing has been measured.

## Part A -- Reading and formatting data using R

### 1. Read data in R

Reading functions in *R* follows common rules with similar arguments. The most important are :

-   *name=* for the file name you need to read (with the path if not in your working directory)
-   *header=* to specify if your file contains a header with column names
-   *sep=* to specificy the type of column separator (",", ";", "/")

The `read.csv()` function is a generic function to read tabulated data set. To store the output of the `read.csv()` function in a R object we use the assignment operator, `<-`. For example, here we name our object *sleepapnea*

Click *Run Code* to give it a try:

```{r read, exercise = TRUE, exercise.eval = FALSE}
sleepapnea <- read.csv("data/SleepApnea.csv", header=TRUE, sep=",")
```

For a quick verification of the reading output, you can use `head()` or `tail()` to observe the first 6 or last 6 rows for your R object *sleepapnea*.

Run the command below to test the `head()` function:

```{r header, exercise = TRUE, exercise.eval = FALSE}
head(sleepapnea)
```

Run the command in the code chunk below to test the `tail()` function

```{r tail, exercise = TRUE, exercise.eval = FALSE}
tail(sleepapnea)
```

Try to modify the arguments of the `tail()` function to display the last 10 rows of your object (If you need help, use `?tail` in the code chunk)

```{r tailn10, exercise = TRUE}
```

```{r tailn10-solution}
tail(sleepapnea, n=10)
```

**Good job!** You should now have a sense of :

-   what is a R function,
-   what are arguments for a function,
-   how to the store the result of your function call using the assignment operator, `<-`, so you keep and can reuse the result later on.

### 2. Data structure and variable types

To Look at your data structure in the R console you can use the `str()` function.

```{r str, exercise = TRUE, exercise.eval = FALSE}
str(sleepapnea)
```

Your data set is a `data.frame` which is a rectangular R data format where variables can be of different types (e.g. numeric, character, factor...)

**Looking at your object, is the output consistent with what you expected in terms of variables' types?**

Note: The output of `str()` is equivalent to the visual display of the blue dot with the white arrow tag to you object's name in the the *Environment panel* of RStudio

### 3. Data type conversion

The *number* variable in the sleep apnea data set corresponds to the identifier of each patient. It should not be a numeric value but a character or factor (categorical variable).

a.  First, we transform into character the *number* variable using the function `as.character()`. One way of doing is to overwrite the *number* variable by selecting the vector(variable) *number* in the data set using the `$` sign and applying the `as.character()`.

```{r tocharacter, exercise = TRUE, exercise.eval = FALSE}
sleepapnea$number <- as.character(sleepapnea$number)
```

Similarly, the *gender* variable should be a factor (categorical variable) and not a character.

b.  Second, we need to transform the *gender* variable using `factor()`

```{r tofactor, exercise = TRUE, exercise.eval = FALSE}
summary(sleepapnea$gender)
sleepapnea$gender <- factor(sleepapnea$gender)
summary(sleepapnea$gender)
```

Note that we propose to use the summary function before and after to assess the result of applying the function.

c.  Similarly, we need to transform the *diabetes* variable (read as a numeric) into a categorical variable using the function `factor()` and setting the argument `labels=` to "No" and "Yes" to annotate the levels 0 and 1

```{r tofactorwithlabels, exercise = TRUE, exercise.eval = FALSE}
summary(sleepapnea$diabetes)
sleepapnea$diabetes <- factor(sleepapnea$diabetes, levels = c(0,1),
                              labels=c("No","Yes"))
summary(sleepapnea$diabetes)
```

d.  To practice, do the same with the *Trigly_4Classes* variable knowing that :

-   classe 1 is $< 1.50$ or Normal rate
-   classe 2 is $1.50-1.99$ or Borderline rate
-   classe 3 is $2.00-4.99$ or High rate
-   classe 4 is $> 5.00$ or Very high rate

```{r TriglycerideCl, exercise = TRUE}
```

```{r TriglycerideCl-solution}
sleepapnea$Triglyceride_Cl_Ref <- factor(sleepapnea$Triglyceride_Cl_Ref, levels = c(1:4),
                              labels=c("Normal","Bordeline",
                                       "High", "Very High"))
```

### 4. Save the clean data set

To keep the results of your pre-processing and data management steps, it is recommended that you use the `save()` function to create a *Rdata file* that contains the clean version of your data set.

```{r save, exercise = TRUE, exercise.eval = FALSE}
sleepapnea_clean <- sleepapnea
save(sleepapnea_clean, file="sleepapnea_clean.Rdata")
```

## Part B - Descriptive Analysis

### 1. Load your clean data set

```{r loadcache, echo=FALSE}
load("../../data/sleepapnea_clean.Rdata")
```

```{r load, exercise = TRUE, exercise.eval = FALSE}
load("sleepapnea_clean.Rdata")
```

### 2. The summary() and table() functions

a.  Describe the variable gender using the functions `summary()` and `table()`

```{r gendersummary, exercise = TRUE}
```

```{r gendersummary-solution}
summary(sleepapnea_clean$gender)
table(sleepapnea_clean$gender)
```

b.  Compute contingency table between *gender* and *diabetes* using the function `table()`

```{r gender2diabetes, exercise = TRUE}
```

```{r gender2diabetes-solution}
table("gender"=sleepapnea_clean$gender, "diabetes"=sleepapnea_clean$diabetes)
```

Using `prop.table()` to compute the proportion of diabetes among female and male

```{r propgender2diabetes, exercise = TRUE}
```

```{r propgender2diabetes-solution}
round(prop.table(table("gender"=sleepapnea_clean$gender, "diabetes"=sleepapnea_clean$diabetes), margin=1)*100, 2)
```

c.  Describe the variable age using the function `summary()`

```{r agesummary, exercise = TRUE}
```

```{r agesummary-solution}
summary(sleepapnea_clean$age)
```

d.  Compute the mean, standard deviation, mediane, quartiles for the variable age

```{r agedesc, exercise = TRUE}
```

```{r agedesc-solution}
mean(sleepapnea_clean$age, na.rm=TRUE)
sd(sleepapnea_clean$age, na.rm=TRUE)
median(sleepapnea_clean$age, na.rm=TRUE)
quantile(sleepapnea_clean$age, na.rm=TRUE)
quantile(sleepapnea_clean$age, probs= c(0, 0.3, 0.6), na.rm=TRUE)
```

e.  What is the best graph for the variable age?

```{r agehistogram}
hist(sleepapnea_clean$age, main="distribution of age", xlab="age (years)")
```

Draw the age distribution using the hist() function. Choose a width of age class of 2 years and of 5 years. Discuss about the shape of the two distributions.

```{r agehist2, exercise = TRUE}
```

```{r agehist2-solution}
hist(sleepapnea_clean$age, width=2, main="distribution of age", xlab="age (years)")
hist(sleepapnea_clean$age, width=5, main="distribution of age", xlab="age (years)")
```

f.  The `boxplot()` function summarize the quantile parameters for a quantitative variable. Below if the boxplot for the *age* variable. Interpret it.

```{r ageboxplot}
boxplot(sleepapnea_clean$age, main="distribution of age", xlab="age (years)")
```

Draw the box plot of age by gender group. Discuss about the shape of the two distributions.

```{r agebysexboxplot, exercise = TRUE}
```

```{r agebysexboxplot-solution}
boxplot(sleepapnea_clean$age ~  sleepapnea_clean$gender,  main="distribution of age by gender group", ylab="age (years)", xlab="gender")
```

g.  Compute the confidence interval of the age variable knowing the formula follows the Normal law assumption:

-   $1.96*\sigma/\sqrt{n}$

```{r agebyci}
bound <- 1.96*sd(sleepapnea_clean$age)/sqrt(length(sleepapnea_clean$age))
# upper bound
mean(sleepapnea_clean$age)+bound 
# lower bound
mean(sleepapnea_clean$age)-bound 
```

Note: The *epiDisplay* library propose the `ci()` function that computes the confidence interval. To use the function do not forget to load the library in your working session if not already done so.

## Part C - Univariate analysis

### 1. Summary by strata (group)

Is there an association between sleep apnea and body mass index (bmi) status?

a.  First, we need to create the bmi groups. We propose you to `cut()` the bmi (quantitative) variable into 2 groups: $<25$ for "normal" and $>25$ for "overweight"

```{r cut, exercise = TRUE, exercise.eval = FALSE}
sleepapnea_clean$bmi_2grp <- cut(sleepapnea_clean$bmi, c(0, 25, 55), include.lowest = TRUE, right=FALSE)
levels(sleepapnea_clean$bmi_2grp) <- c("normal", "overweight")
```

**Practice:** We propose you to `cut()` the bmi (quantitative) variable into 3 groups: $<25$ for "normal", $[25 to 30[$ for "overweight", and $>30$ for "obese"

```{r, bmi3grp, exercise = TRUE}
```

```{r bmi3grp-solution, exercise = TRUE}
sleepapnea_clean$bmi_3grp <- cut(sleepapnea_clean$bmi, c(0, 25, 30, 55), include.lowest = TRUE, right=FALSE)
levels(sleepapnea_clean$bmi_3grp) <- c("normal", "overweight", "obese")
```

b.  Second, we want to summarize the average of sleep apnea level by the 2 groups of bmi.

```{r meanbygroup, exercise = TRUE, exercise.eval = FALSE}
library(dplyr)
sleepapnea_clean %>% group_by(bmi_2grp) %>% summarise("mean_apnea"=mean(SleepApnea, na.rm = TRUE))
```

**Practice:** Repeat the computation for the bmi in 3 groups

```{r, meanbygroup3, exercise = TRUE}
```

```{r meanbygroup3-solution, exercise = TRUE}
sleepapnea_clean %>% group_by(bmi_3grp) %>% summarise("mean_apnea"=mean(SleepApnea, na.rm = TRUE))
```

**Practice:** Try with other quatitative variables of the data set such as age, height or weight which could be some confounding factors

### 2. Statistical tests of association

**Example 1:** We would like to compare of the mean of sleep apnea among the 2 groups of bmi. What statistical tests to choose? What are the assumptions?

a.  Since we consider the mean we suppose that we have normally distributed values of sleep apnea in the 2 bmi groups. Let's verify using the classical Shapiro test:

```{r apneabmi, exercise = TRUE, exercise.eval = FALSE}
by(sleepapnea_clean$SleepApnea, sleepapnea_clean$bmi_2grp, shapiro.test)
```

What are your conclusions?

b.  Given the result of the Shapiro test, we choose to use the non-parametric test `wilcox.text()` to assess for a potential association. Try to apply the test and interpret the results.

```{r wilcox, exercise = TRUE, exercise.eval = FALSE}
wilcox.test(SleepApnea ~ bmi_2grp, data= sleepapnea_clean)
```

Note: For most statistical tests of association, the *R* syntax is of type formula i.e. `func(y ~ x, data=mydata)` but you might want to look at the help pages of the different test functions for the details of their usage.

c.  Given the result of the Shapiro test, we can try to transform the sleep apnea measures into a logarithm scale to approach a Normal distribution and use the parametric Student T test using `t.test()`. What is your interpretation?

```{r apnealog,  exercise = TRUE}

```

```{r apnealog-solution, exercise = TRUE, exercise.eval = FALSE}
sleepapnea_clean$SleepApnea_log <- log(SleepApnea)
t.test(SleepApnea_log ~ bmi_2grp, data=sleepapnea_clean)
```

**Example 2:** We would like to compare of the distribution of sleep apnea measures among the 3 groups of bmi. Given the repartition of the patients among the 3 groups of bmi, we propose to use the `kruskal.test()` non-parametric rank sum test to assess the differences (or not) between groups.

```{r kruskal, exercise = TRUE, exercise.eval = FALSE}
kruskal.test(SleepApnea ~ bmi_3grp, data= sleepapnea_clean)
```

What is your interpretation?

To identify which groups differ from the other(s) you can use the `dunn_test()` from the `rstatix` library

```{r dunn_test, exercise = TRUE, exercise.eval = FALSE}
library(rstatix)
dunn_test(SleepApnea ~ bmi_3grp, data= sleepapnea_clean)
```

**Example 3:** We would like to test whether there is an association between gender and bmi groups. What statistical tests to choose? What are the assumptions?

a.  We consider the gender and bmi groups which are categorical variables. We can perform a Fisher Exact test using `fisher.test()`

```{r fisher}
fisher.test(sleepapnea_clean$gender, sleepapnea_clean$bmi_2grp)
```

b.  A similar test is the Chi Square test that you can apply using `chisq.test()`. Have a try and interpret the results.

```{r chi2, exercise = TRUE}

```

```{r chi2-solution, exercise = TRUE}
chisq.test(sleepapnea_clean$gender, sleepapnea_clean$bmi_2grp)
```
